---
title: "Reproducible Research: Peer Assessment 1"
author: "TZiegler"
date: "16. Dezember 2015"
output: 
  html_document:
    keep_md: true
---

```{r}
library(knitr)
```



## Loading and preprocessing the data
This step extracts the given zip file, loads the csv file, checks it and converts the date data into real date format for further analysis.

```{r}
## Exctract zipfile
unzip("activity.zip")

activity <- read.csv("activity.csv")
```

Get an owverview of the data.
```{r}
## Show structure of data frame
str(activity)
## Show summary
summary(activity)
```

Convert date values to Date format
```{r}
activity$date <- as.Date(activity$date)
```



## What is mean total number of steps taken per day?
For this part the missing values in the dataset can be ignored.

1. Calculate the total number of steps taken per day.
```{r dailysteps, fig.height=6}
dailysteps <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm = TRUE)

## Replace generic column names with new ones
names(dailysteps) <- c('day', 'totalsteps')

kable(dailysteps, format = "pandoc", col.names = c("Date","Total steps"), format.args = list(big.mark = ','))
```


2. Make a histogram of the total number of steps taken each day.
```{r histogram, fig.height=6}
h <- hist(dailysteps$totalsteps, breaks = 15, main = "Total number of steps per day", col = "red", xlab = "Steps per day (frequency counts on top of each cell)")

## Add the frequency counts above each cell
text(h$mids,h$counts,labels=h$counts, adj=c(0.5, -0.2))
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
steps_mean <- round(mean(dailysteps$totalsteps,na.rm=TRUE),2)
steps_median <- median(dailysteps$totalsteps, na.rm = TRUE)
```
- The mean of the total number of steps taken per day is **`r steps_mean`**
- The median of the total number of steps taken per day is **`r steps_median`**



## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r activitypattern}
avsteps <- tapply(activity$steps, activity$interval, FUN=mean, na.rm = TRUE)
plot(avsteps, type='l', col='red',main="Average daily activity pattern",xlab="5 minutes intervals",ylab="Average numbers of steps")
```
A day has 288 of 5-minute intervals. 

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval <- which(avsteps %in% max(avsteps))
```
The 5-minute interval with the maximum number of steps is interval **`r max_interval`** with an average number of  `r round(max(avsteps), 2)` steps.


## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}

```



## Are there differences in activity patterns between weekdays and weekends?
